---
title: >
  How to use DeeDeeExperiment with single-cell data
author:
- name: Najla Abassi
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  email: abassina@uni-mainz.de
- name: Lea Rothörl
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
- name: Federico Marini
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  - Research Center for Immunotherapy (FZI), Mainz
  email: marinif@uni-mainz.de
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('DeeDeeExperiment')`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    code_folding: show
    code_download: yes
vignette: >
  %\VignetteIndexEntry{2. How to use DeeDeeExperiment with single-cell data}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{DeeDeeExperiment}
  %\VignetteKeywords{GeneExpression, RNASeq, Sequencing, Pathways, Infrastructure}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: DeeDeeExperiment_bibliography.bib
---

<style type="text/css">
.smaller {
  font-size: 10px
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  error    = FALSE,
  warning  = FALSE,
  eval     = TRUE,
  message  = FALSE
)
```

# Differential State (DS) analysis with `muscat` & `DeeDeeExperiment`

## `DeeDeeExperiment` on the `Kang` dataset

In this vignette, we illustrate how to integrate Differential Expression
Analysis (DEA) results generated with the `r BiocStyle::Biocpkg("muscat")`
framework into a `r BiocStyle::Biocpkg("DeeDeeExperiment")` object. As an example,
we use a publicly available dataset from Kang, et al. "Multiplexed droplet single-cell RNA-sequencing using natural genetic variation", published in Nature
Biotechnology, December 2017
[@Kang2017](https://doi.org/10.1038/nbt.4042)

The data is made available via the `r BiocStyle::Biocpkg("ExperimentHub")`
Bioconductor package as a `r BiocStyle::Biocpkg("SingleCellExperiment")` object
containing scRNA-seq data from PBMCs obtained from 8 lupus patients before and
after IFNβ stimulation.

For demonstration purposes we adapt parts of the code from the original `r BiocStyle::Biocpkg("muscat")` [vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/muscat/inst/doc/analysis.html#differential-state-ds-analysis))

```{r loadlib}
library("DeeDeeExperiment")
library("ExperimentHub")
library("scater")
library("muscat")
library("limma")
```

We begin by creating an `ExperimentHub` instance, which provides access to
curated datasets stored in the Bioconductor cloud. Using `query()`, we filter
available records for entries matching the keyword "Kang", and then load the
dataset of interest using its accession ID "EH2259".

```{r load_sce}
# retrieve the data
eh <- ExperimentHub()
query(eh, "Kang")
sce <- eh[["EH2259"]]
```

Before running muscat, we perform standard single-cell preprocessing steps:
removing undetected genes, filtering low-quality cells using `r BiocStyle::Biocpkg("scater")`,
and removing lowly expressed genes.

```{r rm_undetected_genes}
# remove undetected genes
sce <- sce[rowSums(counts(sce) > 0) > 0, ]
```

```{r qc}
# calculate per-cell quality control (QC) metrics
qc <- perCellQCMetrics(sce)
# remove cells with few or many detected genes
ol <- isOutlier(metric = qc$detected, nmads = 2, log = TRUE)
sce <- sce[, !ol]
dim(sce)
```

```{r rm_low_exp_genes}
# remove lowly expressed genes
sce <- sce[rowSums(counts(sce) > 1) >= 10, ]
dim(sce)
```

```{r normalization}
# compute sum-factors & normalize
sce <- computeLibraryFactors(sce)
sce <- logNormCounts(sce)
```

We then prepare the data for `muscat`. The package expects a certain format of
the input SCE. Specifically, the following cell metadata (colData) columns have
to be provided:

* `sample_id` : unique sample identifiers
* `cluster_id` : subpopulation (cluster) assignments
* `group_id` : experimental group/condition


```{r prep}
# data preparation
sce$id <- paste0(sce$stim, sce$ind)
(sce <- prepSCE(sce,
                kid = "cell", # subpopulation assignments
                gid = "stim",  # group IDs (ctrl/stim)
                sid = "id",   # sample IDs (ctrl/stim.1234)
                drop = TRUE))  # drop all other colData columns
```

We compute UMAP for visualization. The dataset already includes precomputed TSNE
coordinates, so we only run UMAP here.

```{r reddim}
# compute UMAP using 1st 20 PCs
sce <- runUMAP(sce, pca = 20)
```

Then we aggregate measurements for each sample (in each cluster) to obtain pseudobulk data

```{r aggregate_by_cell_type}
# aggregate by cell type
pb <- aggregateData(
  sce,
  assay = "counts",
  fun = "sum",
  by = c("cluster_id", "sample_id")
)

assayNames(pb)
```

And construct the contrast matrix

```{r design}
# construct design & contrast matrix
ei <- metadata(sce)$experiment_info
mm <- model.matrix(~ 0 + ei$group_id)
dimnames(mm) <- list(ei$sample_id, levels(ei$group_id))
contrast <- makeContrasts("stim-ctrl", levels = mm)
```

With the pseudobulk data assembled, we can now test for differential state (DS)
using `pbDS`

```{r pbDS}
# run DS analysis
muscat_res <- pbDS(pb, design = mm, contrast = contrast)

names(muscat_res$table[["stim-ctrl"]])
```

Now we integrate the output of muscat in `muscat_list_for_dde()` to transform it
into a format accepted by `DeeDeeExperiment`

```{r create_muscat_list}

muscat_list <- muscat_list_for_dde(res = list(`stim-ctrl` = muscat_res),
                                   padj_col = "p_adj.loc")
```

Finally the results can be directly added to a new `dde` object, or to an
existing one.

```{r create_dde}
# create dde
dde <- DeeDeeExperiment(sce = sce,
                        de_results = muscat_list)
dde
```

```{r get_DEA_names}
#check DEAs
getDEANames(dde)
```
```{r get_DEAs}
# retrieve results
getDEA(dde,
       dea_name = "stim-ctrl_NK cells")|> head()

getDEA(dde,
    dea_name = "stim-ctrl_CD14+ Monocytes",
    format = "original"
) |> head()
```

```{r get_DEA_info}
dea_name <- "stim-ctrl_B cells"
getDEAInfo(dde)[[dea_name]][["package"]]
getDEAInfo(dde)[[dea_name]][["package_version"]]
```

```{r add_scenario}
dde <- addScenarioInfo(dde,
                       dea_name = "stim-ctrl_Dendritic cells",
                       info = "This result contains the output of pseudobulk DE analysis performed on dendritic cells, comparing untreated samples to those stimulated with IFNβ")
```

```{r get_summary}
summary(dde, show_scenario_info = TRUE)
```

# Session info {.unnumbered .smaller}

```{r sessioinfo}
sessionInfo()
```

# References {.unnumbered}
